Advancements in Moltres for Time-Dependent Multiphysics Molten Salt Reactor Modeling

### 2024-11-13 transient.tex
Full draft + edits

This work aims to replicate the integral count curve following a rod
drop of Rod 1 with the $^{235}$U loading corresponding to "30 capsule additions" of highly-enriched
$^{235}$U capsules. Each capsule contains 85 g of $^{235}$U, of which 94.94 \% enters the core due
to residual fuel salt remaining in the drain tanks after mixing \cite{fratoni_molten_2020}.

@@

The \gls{MSRE} achieved initial criticality after eight capsule additions with Rod 1 inserted to a
height of 46.6 inches relative to the fully inserted height \cite{prince_zero-power_1968}.
From the \gls{MSRE} experimental rod worth
data, the rod worth evaluates to approximately 72 pcm relative to the fully withdrawn state; the
$k_\text{eff}$ of the \gls{MSRE} at initial criticality with Rod 1 fully withdrawn would be
1.00072. According to the same \gls{MSRE} report, the
authors back-calculated the critical $^{235}$U loading with Rod 1 fully withdrawn to be 65.25 kg.
They also provided a $^{235}$U concentration reactivity coefficient of 0.223 $(\delta k/k)/\delta m
/m)$. The excess $^{235}$U mass at initial criticality compensating for Rod 1 inserted to 46.6
inches can estimated through the following steps:
%
\begin{align}
  \frac{\Delta k / k}{\Delta m / m} &= 0.223 \nonumber \\
  \frac{m}{k}\frac{dk}{dm} &= 0.223 \nonumber \\
  \int^{1.00072}_1 \frac{1}{k}\ dk &= 0.223 \int^{65.25+\Delta m}_{65.25} \frac{1}{m}\ dm \nonumber \\
  \ln 1.00072 &= 0.223 \ln\left(\frac{65.25+\Delta m}{65.25}\right) \nonumber \\
  \Delta m &= 0.2109 \mbox{ kg}.
\end{align}
%
This change in mass is equivalent to approximately 2.61 highly-enriched $^{235}$U capsules.
Therefore, the $^{235}$U loading for the rod drop experiment is:
%
\begin{gather}
  m_\text{rod drop} = 65.25 \text{kg} + (30 - 8 + 2.61) * 0.085 \text{kg} * 0.9494 = 67.24 \text{kg}
\end{gather}
%
The $k_\text{eff}$ of the \gls{MSRE} with 67.24 kg loading and all rods fully withdrawn is:
%
\begin{align}
  \int^{k_{67.24 \text{kg}}}_1 \frac{1}{k}\ dk &= 0.223 \int^{67.24}_{65.25} \frac{1}{m}\ dm \nonumber \\
    k_{67.24 \text{kg}} &= 1.00671
\end{align}

@@

This work aims to replicate the power response following a reactivity insertion of 24.8 pcm of the
\gls{MSRE} initially at 1 MW. Figure \ref{msre-reactivity-insertion} shows the theoretical and
experimental power response following the reactivity insertion. Three separate publications
\cite{zanetti_innovative_2015, singh_study_2019, mochizuki_validation_2022} noted that the plot is
mislabeled with the wrong reactivity insertion value of 0.0139 \% $\delta k/k$, which had been the
reactivity insertion value for the 8 MW case. The correct reactivity insertion value is 24.8 pcm as
mentioned prior.

@@

For the reactivity insertion simulation, we must first identify the $^{233}$U salt composition used
in the experiment. The \gls{MSRE} $^{233}$ power operations occurred during Runs 16 to 18 according
to \gls{MSRE} reports \cite{thoma_chemical_1971, burke_analysis_2019}. Table \ref{table:u233-salt}
lists the proportion of elements in the \gls{MSRE} fuel salt during Runs 16 to 18. Noted that the
components do not sum up to 100 \% due to experimental uncertainty in chemical analysis and the
presence of other elements that were not measured. Taking the nominal uranium value to be true, I scaled
up the compositions of non-uranium elements to make up the remaining $(100\%-0.8155\%)$ of the salt. I
kept the nominal weight percentage of uranium constant because it is the most accurate value in the table
and has the biggest impact on the reactivity.
%
\begin{table}[t]
  \small
  \centering
  \setlength\tabcolsep{4pt}
  \caption{\gls{MSRE} fuel salt mean composition during the power operations with $^{233}$U \cite{thoma_chemical_1971}.
  All values were measured via chemical analysis except the nominal U weight percentage, which was
  calculated from inventory tracking of cumulative U additions to the salt.}
  \begin{tabular}{l S[table-format=2.2(2)] S[table-format=1.2(2)] S[table-format=2.2(2)] S[table-format=2.2(3)] S[table-format=1.3(2)] S}
    \toprule
    Element & {Li} & {Be} & {Zr} & {F} & {U} & {U (Nominal)} \\
    \midrule
    Weight percentage [\%]  & 11.33(39) & 6.71(30) & 10.99(29) & 68.71(126) & 0.809(11) & 0.8155 \\
    \bottomrule
  \end{tabular}
  \begin{tabular}{l S[table-format=2(1)] S[table-format=3(2)] S[table-format=2(2)] S}
    \toprule
    Element & {Cr} & {Fe} & {Ni} & {O} \\
    \midrule
    Parts per million [ppm] & 72(9) & 135(24) & 46(12) & 61 \\
    \bottomrule
  \end{tabular}
  \label{table:u233-salt}
\end{table}

@@

The second time-dependent simulation is based on a reactivity insertion experiment on the
$^{233}$U-fueled \gls{MSRE}. A control rod withdrawal induces a 24.8 pcm step reactivity
insertion on the \gls{MSRE} initially running at 1 MW with steady salt flow. Moltres reproduces
the general shape of the first power peak and the subsequent power oscillations. The power peak
magnitude differs by y \% compared to \gls{MSRE} data. The fuel
salt, graphite, inlet, and outlet temperature measurements show delayed responses to the increase
in power output due to the time taken for the salt and graphite to heat up. Over the simulation
time, the numerical \gls{MSRE} model settles into higher temperature distributions than the initial
state to compensate for the reactivity insertion.

>>>>

This work aims to replicate the integral count curve following a rod drop of Rod 1 with the
$^{235}$U loading corresponding to ``30 capsule additions'' of highly-enriched $^{235}$U capsules
into the ``pump bowl''. Each capsule contains 85 g of $^{235}$U, of which 94.94 \% enters the core
due to residual fuel salt remaining in the drain tanks after mixing \cite{fratoni_molten_2020}.

@@

The \gls{MSRE} achieved initial criticality after eight capsule additions with Rod 1 inserted to a
height of 46.6 inches relative to the fully inserted height \cite{prince_zero-power_1968}.
From the \gls{MSRE} experimental rod worth
data, the rod worth evaluates to approximately 72 pcm relative to the fully withdrawn state; the
$k_\text{eff}$ of the \gls{MSRE} at initial criticality with Rod 1 fully withdrawn would be
1.00072. According to the same \gls{MSRE} report, the
authors back-calculated the critical $^{235}$U loading with Rod 1 fully withdrawn to be 65.25 kg.
They also provided a $^{235}$U concentration reactivity coefficient of 0.223 $(\delta k/k)/\delta m
/m)$. The excess $^{235}$U mass at initial criticality compensating for Rod 1 inserted to 46.6
inches can estimated through the following steps:
%
\begin{align}
  \frac{\Delta k / k}{\Delta m / m} &= 0.223 \nonumber \\
  \frac{m}{k}\frac{dk}{dm} &= 0.223 \nonumber \\
  \int^{1.00072}_1 \frac{1}{k}\ dk &= 0.223 \int^{65.25+\Delta m}_{65.25} \frac{1}{m}\ dm \nonumber \\
  \ln \left(1.00072\right) &= 0.223 \ln\left(\frac{65.25+\Delta m}{65.25}\right) \nonumber \\
  \Delta m &= 0.2109 \mbox{ kg}.
\end{align}
%
This change in mass is equivalent to approximately 2.61 highly-enriched $^{235}$U capsules.
Therefore, the $^{235}$U loading for the rod drop experiment is:
%
\begin{gather}
  m_\text{rod drop} = 65.25 \text{kg} + (30 - 8 + 2.61) * 0.085 \text{kg} * 0.9494 = 67.24 \text{kg}
\end{gather}
%
The $k_\text{eff}$ of the \gls{MSRE} with 67.24 kg loading and all rods fully withdrawn is:
%
\begin{align}
  \int^{k_{67.24 \text{kg}}}_1 \frac{1}{k}\ dk &= 0.223 \int^{67.24}_{65.25} \frac{1}{m}\ dm \nonumber \\
    k_{67.24 \text{kg}} &= 1.00671
\end{align}

@@

This work aims to replicate the power response following a reactivity insertion of 24.8 pcm of the
\gls{MSRE} initially at 1 MW. Figure \ref{fig:msre-reactivity-insertion} shows the theoretical and
experimental power response following the reactivity insertion. Three separate publications
\cite{zanetti_innovative_2015, singh_study_2019, mochizuki_validation_2022} noted that the plot is
mislabeled with the wrong reactivity insertion value of 0.0139 \% $\delta k/k$, which had been the
reactivity insertion value for the 8 MW case. The correct reactivity insertion value is 24.8 pcm as
mentioned prior.

@@

For the reactivity insertion simulation, we must first identify the $^{233}$U salt composition used
in the experiment. Power operations with the $^{233}$U-fueled \gls{MSRE} core
occurred during Runs 16 to 18 according
to \gls{MSRE} reports \cite{thoma_chemical_1971, burke_analysis_2019}. Table \ref{table:u233-salt}
lists the proportion of elements in the \gls{MSRE} fuel salt during Runs 16 to 18. Note that the
components do not sum up to 100 \% due to experimental uncertainty in chemical analysis and the
presence of other elements that were not measured. Taking the nominal uranium value to be true, I scaled
up the compositions of non-uranium elements to make up the remaining $(100\%-0.8155\%)$ of the salt. I
kept the nominal weight percentage of uranium constant because it is the most accurate value in the table
and has the biggest impact on the reactivity.
%
\begin{table}[t]
  \small
  \centering
  \setlength\tabcolsep{4pt}
  \caption{\gls{MSRE} fuel salt mean composition during the power operations with $^{233}$U \cite{thoma_chemical_1971}.
  All values were measured via chemical analysis except the nominal U weight percentage, which was
  calculated from inventory tracking of cumulative U additions to the salt.}
  \begin{tabular}{l S[table-format=2.2(2)] S[table-format=1.2(2)] S[table-format=2.2(2)] S[table-format=2.2(3)] S[table-format=1.3(2)] S}
    \toprule
    Element & {Li} & {Be} & {Zr} & {F} & {U} & {U (Nominal)} \\
    \midrule
    Weight percentage [\%]  & 11.33(39) & 6.71(30) & 10.99(29) & 68.71(126) & 0.809(11) & 0.8155 \\
    \bottomrule
  \end{tabular}
  \begin{tabular}{l S[table-format=2(1)] S[table-format=3(2)] S[table-format=2(2)] S}
    \toprule
    Element & {Cr} & {Fe} & {Ni} & {O} \\
    \midrule
    Parts per million [ppm] & 72(9) & 135(24) & 46(12) & 61 \\
    \bottomrule
  \end{tabular}
  \label{table:u233-salt}
\end{table}

@@

The second time-dependent simulation is based on a reactivity insertion experiment on the
$^{233}$U-fueled \gls{MSRE}. A control rod withdrawal induces a 24.8 pcm step reactivity
insertion on the \gls{MSRE} initially running at 1 MW with steady salt flow. Moltres reproduces
the general shape of the first power peak and the subsequent power oscillations. The power peak
magnitude differs by y \% compared to \gls{MSRE} data. The fuel
salt, graphite, inlet, and outlet temperature measurements show delayed responses to the increase
in power output due to the time taken for the salt and graphite to heat up. The temperature values
eventually settle higher than the initial
state to compensate for the reactivity insertion.
